Supplementary materials for: Modelling palaeoecological time series using generalized additive models

### 2023-10-16 supplementary-materials.Rmd
update supplementary materials to work with current gratia

---
title: "Supplementary materials for: Modelling palaeoecological time series using generalized additive models"
author: Gavin L. Simpson
affiliation: "Institute of Environmental Change and Society, \\ University of Regina, Regina, Saskatchewan, Canada, S4S 0A2"
mail: "gavin.simpson@uregina.ca"
date: August 16, 2018
keywords: time series; generalized additive model; simultaneous interval; spline; environmental change 
bibliography: references.bib
csl: frontiers-in-ecology-and-evolution.csl
geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
fontsize: 12pt
mainfont: TeX Gyre Pagella
output:
  pdf_document:
    highlight: tango
    number_sections: true
    fig_width: 8
    fig_height: 5
    fig_crop: false
    keep_tex: true
    latex_engine: xelatex
    md_extensions: +header_attributes+superscript+subscript
    includes:
      in_header: header.tex
  md_document:
    variant: markdown_github
---

@@

The following packages are required: *mgcv*, *scam*, *ggplot2*, *cowplot*, and *tidyr*.

```{r load-packages, cache = FALSE}
library("mgcv")
library("scam")
library("ggplot2")
library("cowplot")
library("tidyr")
```
In addition, the *gratia* package is required; this in-development package is not on CRAN but can be installed directly from GitHub using functions from the *devtools* package. To install the package, install *devtools* and then use `devtools::install_github()` to install *gratia*, as shown below:

```{r install-gratia, eval = FALSE}
## gratia is not on CRAN, install from github:
install.packages("devtools")
devtools::install_github("gavinsimpson/gratia")
```
Assuming that the installation of *gratia* completes without error, the package can be loaded as usual
```{r load-gratia, cache = FALSE}
library("gratia") # need to change the name of the package
```

@@

car1Plt <- ggplot(car1, aes(x = S, y = Correlation)) +
    geom_ribbon(aes(ymax = Upper, ymin = Lower),
                fill = "black", alpha = 0.2) +
    geom_line() +
    ylab(expression(italic(h) * (list(Delta[t], varphi)))) +
    xlab(expression(Delta[t] ~ (years)))
car1Plt
```
The exponential decline in correlation with increasing separation is evident here; once samples are ~10 years apart, there is little estimated dependence between them.

@@

The same model as used for Small Water was also initially attempted with the Braya-Sø data. However, in order to even fit the model with both a smooth and the CAR(1) process, I had to change the default optimizer used to estimate the model parameters, and reduce the basis dimension, `k`, to a small number. I also fit the model using GCV, which is, at the time of writing, the default in `gam()`, hence no `method` argument

```{r fit-braya-so-car1-and-gcv-models}
## fit the car(1) model --- needs optim as this is not a stable fit!
## also needs k setting lower than default
braya.car1 <- gamm(UK37 ~ s(Year, k = 5), data = braya, 
                   correlation = corCAR1(form = ~ Year),
                   method = "REML",
		           control = list(niterEM = 0, optimMethod = "BFGS", 
                                  opt = "optim"))

@@

## do the simulations
sims <- simulate(mod, nsim = nsim, newdata = newYear, unconditional = TRUE)

## rearrange the output into a long/tidy format
colnames(sims) <- paste0("sim", seq_len(nsim))
sims <- setNames(stack(as.data.frame(sims)), c("simulated", "run"))
sims <- transform(sims, Year = rep(newYear$Year, nsim),
                  simulated = simulated)

## Plot simulated trends
smallSim.plt <- ggplot(newYear, aes(x = Year, y = fit)) +
    geom_line(data = sims,
              mapping = aes(y = simulated, x = Year, group = run),
              colour = "grey80") +
    geom_line(lwd = 1) +
    labs(y = d15n_label, x = "Year CE")
smallSim.plt
```

@@

## simulate
set.seed(1)
sims2 <- simulate(braya_reml, nsim = nsim, newdata = newBraya,
                  unconditional = TRUE)

## rearrange the output into a long/tidy format
colnames(sims2) <- paste0("sim", seq_len(nsim))
sims2 <- setNames(stack(as.data.frame(sims2)),
                  c("simulated", "run"))
sims2 <- transform(sims2, Year = rep(newBraya$Year, nsim),
                   simulated = simulated)

brayaSim.plt <- ggplot(braya_pred, aes(x = Year, y = fit)) +
    geom_line(data = sims2,
              mapping = aes(y = simulated, x = Year, group = run),
              colour = "grey80") +
    geom_line(lwd = 1) +
    labs(y = braya_ylabel, x = "Year CE")
brayaSim.plt
```

@@

```{r small-compare-intervals}
sw.cint <- confint(mod, parm = "Year", newdata = newYear,
                   type = "confidence")
sw.sint <- confint(mod, parm = "Year", newdata = newYear,
                   type = "simultaneous")
head(sw.sint)
```

@@

```{r plot-small-water-intervals}
smallInt.plt <- ggplot(sw.cint, aes(x = Year, y = est)) +
    geom_ribbon(data = sw.sint,
                mapping = aes(ymin = lower, ymax = upper, x = Year),
                fill = "grey80", inherit.aes = FALSE) +
    geom_ribbon(mapping = aes(ymin = lower, ymax = upper, x = Year),
                fill = "grey60", inherit.aes = FALSE) +
    geom_line(lwd = 1) +
    labs(y = d15n_label, x = "Year CE")
smallInt.plt
```

The intervals for Braya-Sø are created
```{r braya-compare-intervals}
bs.cint <- confint(braya_reml, parm = "Year", newdata = newBraya,
                   type = "confidence")
bs.sint <- confint(braya_reml, parm = "Year", newdata = newBraya,
                   type = "simultaneous")
```
and plotted
```{r plot-braya-so-intervals}
brayaInt.plt <- ggplot(bs.cint, aes(x = Year, y = est)) +
    geom_ribbon(data = bs.sint,
                mapping = aes(ymin = lower, ymax = upper, x = Year),
                fill = "grey80", inherit.aes = FALSE) +
    geom_ribbon(mapping = aes(ymin = lower, ymax = upper, x = Year),
                fill = "grey60", inherit.aes = FALSE) +
    geom_line(lwd = 1) +
    labs(y = braya_ylabel, x = "Year CE")
brayaInt.plt
```
in the same way

@@

```{r}
## monotonic spline age-depth model
swAge$Error[1] <- 1.1
swAgeMod <- scam(Date ~ s(Depth, k = 5, bs = "mpd"), data = swAge,
                 weights = 1 / swAge$Error, gamma = 1.4)
```

@@

Next, predict from the estimated age model, and draw 25 samples from the posterior distribution using `simulate()`. The results are tidied into a format suitable for further processing and plotting. Note that the posterior samples here are only used for plotting.

```{r, dependson = -1}
## predict from the age model for a smooth set of points in `Depth`
newAge <- with(swAge, data.frame(Depth = seq(min(Depth), max(Depth),
                                             length.out = 200)))
newAge <- transform(newAge,
                    fitted = predict(swAgeMod, newdata = newAge, 
                                     type = "response"))
newSims <- as.data.frame(simulate(swAgeMod, nsim = 25, newdata = newAge))
newSims <- cbind(Depth = newAge$Depth, newSims)
newSims <- gather(newSims, Simulation, Age, -Depth)
```

In the next code chunk, I draw 100 samples from the posterior distribution of the age model, but notice that I pass in the `small` data to `newdata` in the call to `simulate()` as the locations I want new age estimates for are the depths for which we have δ^15^N values. A small function (`fitSWModels`) is written to prepare each simulation for fitting and then actually fit the GAM plus CAR(1) model using the updated age information.

```{r, dependson = -1}
## simulate from age model; each column is a simulation
ageSims <- simulate(swAgeMod, nsim = 100, newdata = small, seed = 42)
ageSims <- as.data.frame(ageSims)

fitSWModels <- function(x, y, knots) {
    dat <- data.frame(d15N = y, Year = x)
    m <- gamm(d15N ~ s(Year, k = 15), data = dat, method = "REML",
              correlation = corCAR1(form = ~ Year), knots = knots)
}

## generate new trends using draws from age-model posterior
simTrendMods <- lapply(ageSims, fitSWModels, y = small$d15N, knots = knots)

@@

```{r, dependson = -1}
## wrapper to simulate from a fitted GAM with the required arguments
simulateSWModels <- function(mod, newdata, nsim, seed = 42) {
    sims <- simulate(mod, nsim = nsim, newdata = newdata, seed = seed)
    as.vector(sims)
}

@@

```{r small-scam-fit-plots, dependson = -1}
## plot the estimated age model ad posterior simulations from it
plt1 <- ggplot(swAge, aes(y = Date, x = Depth)) +
    geom_line(data = newSims,
              mapping = aes(y = Age, x = Depth, group = Simulation),
              alpha = 1, colour = "grey80") +
    geom_line(data = newAge, mapping = aes(y = fitted, x = Depth)) +
    geom_point(size = 1.5, colour = "red") +
    geom_errorbar(aes(ymin = Date - Error, ymax = Date + Error, width = 0),
                  colour = "red") +
    labs(y = "Year CE", x = "Depth (cm)")

>>>>

---
title: "Supplementary materials for: Modelling palaeoecological time series using generalized additive models"
author: Gavin L. Simpson
affiliation: "Institute of Environmental Change and Society, \\ University of Regina, Regina, Saskatchewan, Canada, S4S 0A2"
mail: "gavin.simpson@uregina.ca"
date: October 16, 2023
keywords: time series; generalized additive model; simultaneous interval; spline; environmental change 
bibliography: references.bib
csl: frontiers-in-ecology-and-evolution.csl
geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
fontsize: 12pt
mainfont: TeX Gyre Pagella
output:
  pdf_document:
    highlight: tango
    number_sections: true
    fig_width: 8
    fig_height: 5
    fig_crop: false
    keep_tex: true
    latex_engine: xelatex
    md_extensions: +header_attributes+superscript+subscript
    includes:
      in_header: header.tex
  md_document:
    variant: markdown_github
---

@@

The following packages are required: *mgcv*, *scam*, *ggplot2*, *cowplot*, *dplyr*, and *tidyr*.

```{r load-packages, cache = FALSE}
library("mgcv")
library("scam")
library("ggplot2")
library("cowplot")
library("tidyr")
library("dplyr")
```
In addition, the *gratia* package is required; install *gratia*, using:

```{r install-gratia, eval = FALSE}
## requires development version of gratia that is not on CRAN, install from github:
install.packages("gratia", repos = c("https://gavinsimpson.r-universe.dev",
    "https://cloud.r-project.org"))
stopifnot(packageVersion("gratia") >= "0.8.1.42")
```
Assuming that the installation of *gratia* completes without error, the package can be loaded as usual
```{r load-gratia, cache = FALSE}
library("gratia") # need to change the name of the package
```

@@

car1Plt <- ggplot(car1, aes(x = S, y = Correlation)) +
    geom_ribbon(aes(ymax = Upper, ymin = Lower),
                fill = "black", alpha = 0.2) +
    geom_line() +
    ylab(expression(italic(c) * (list(h, varphi)))) +
    xlab(expression(h ~ (years)))
car1Plt
```
The exponential decline in correlation with increasing separation is evident here; once samples are ~10 years apart, there is little estimated dependence between them.

@@

The same model as used for Small Water was also initially attempted with the Braya-Sø data. I also fit the model using GCV, which is, at the time of writing, the default in `gam()`, hence no `method` argument

```{r fit-braya-so-car1-and-gcv-models, error = TRUE}
## fit the car(1) model --- needs optim as this is not a stable fit!
## also needs k setting lower than default
braya.car1 <- gamm(UK37 ~ s(Year, k = 10), data = braya,
                   correlation = corCAR1(form = ~ Year),
                   method = "REML")

@@

## do the simulations
sims <- smooth_samples(mod$gam, n = nsim, data = newYear,
    unconditional = TRUE)
sims <- mutate(sims, value = value + coef(mod$gam)[1L]) # add intercept

## Plot simulated trends
smallSim.plt <- ggplot(newYear, aes(x = Year, y = fit)) +
    geom_line(data = sims,
              mapping = aes(y = value, x = Year, group = draw),
              colour = "grey80") +
    geom_line(lwd = 1) +
    labs(y = d15n_label, x = "Year CE")
smallSim.plt
```

@@

## simulate
set.seed(1)
sims2 <- smooth_samples(braya_reml, n = nsim, data = newBraya,
    unconditional = TRUE)
sims2 <- mutate(sims2, value = value + coef(braya_reml)[1L]) # add intercept

brayaSim.plt <- ggplot(braya_pred, aes(x = Year, y = fit)) +
    geom_line(data = sims2,
              mapping = aes(y = value, x = Year, group = draw),
              colour = "grey80") +
    geom_line(lwd = 1) +
    labs(y = braya_ylabel, x = "Year CE")
brayaSim.plt
```

@@

```{r small-compare-intervals}
sw.cint <- confint(mod, parm = "s(Year)", data = newYear, type = "confidence")
sw.sint <- confint(mod, parm = "s(Year)", data = newYear, type = "simultaneous")
sw.sint
```

@@

```{r plot-small-water-intervals}
smallInt.plt <- ggplot(sw.cint, aes(x = Year, y = .estimate)) +
    geom_ribbon(data = sw.sint,
                mapping = aes(ymin = .lower_ci, ymax = .upper_ci, x = Year),
                fill = "grey80", inherit.aes = FALSE) +
    geom_ribbon(mapping = aes(ymin = .lower_ci, ymax = .upper_ci, x = Year),
                fill = "grey60", inherit.aes = FALSE) +
    geom_line(lwd = 1) +
    labs(y = d15n_label, x = "Year CE")
smallInt.plt
```

The intervals for Braya-Sø are created
```{r braya-compare-intervals}
bs.cint <- confint(braya_reml, parm = "s(Year)", data = newBraya,
                   type = "confidence")
bs.sint <- confint(braya_reml, parm = "s(Year)", data = newBraya,
                   type = "simultaneous")
```
and plotted
```{r plot-braya-so-intervals}
brayaInt.plt <- ggplot(bs.cint, aes(x = Year, y = .estimate)) +
    geom_ribbon(data = bs.sint,
                mapping = aes(ymin = .lower_ci, ymax = .upper_ci, x = Year),
                fill = "grey80", inherit.aes = FALSE) +
    geom_ribbon(mapping = aes(ymin = .lower_ci, ymax = .upper_ci, x = Year),
                fill = "grey60", inherit.aes = FALSE) +
    geom_line(lwd = 1) +
    labs(y = braya_ylabel, x = "Year CE")
brayaInt.plt
```
in the same way

@@

```{r}
## monotonic spline age-depth model
swAge$Error[1] <- 1
swAgeMod <- scam(Date ~ s(Depth, k = 5, bs = "cv"), data = swAge,
    weights = (1 / swAge$Error) / mean(1 / swAge$Error), gamma = 1)
```

@@

Next, predict from the estimated age model, and draw 25 samples from the posterior distribution using `smooth_samples()`. Note that the posterior samples here are only used for plotting.

```{r, dependson = -1}
## predict from the age model for a smooth set of points in `Depth`
newAge <- data_slice(swAgeMod, Depth = evenly(Depth, n = 200))
newAgeFit <- fitted_values(swAgeMod, data = newAge)
newSims <- smooth_samples(swAgeMod, n = 25, data = newAge) %>%
    mutate(value = value + coef(swAgeMod)[1L]) # %>%
    #left_join(mutate(newAge, .row = row_number()), by = join_by(.row == .row))
```

In the next code chunk, I draw 100 samples from the posterior distribution of the age model, but notice that I pass in the `small` data to `data` in the call to `smooth_samples()` as the locations I want new age estimates for the depths for which we have δ^15^N values. A small function (`fitSWModels`) is written to prepare each simulation for fitting and then actually fit the GAM plus CAR(1) model using the updated age information.

```{r, dependson = -1}
## simulate from age model; each column is a simulation
ageSims <- smooth_samples(swAgeMod, n = 100, data = small, seed = 42)

fitSWModels <- function(x, y, knots) {
    dat <- data.frame(d15N = y,
        Year = pull(x, value) + coef(swAgeMod, parameterized = TRUE)[1L])
    m <- gamm(d15N ~ s(Year, k = 15), data = dat, method = "REML",
              correlation = corCAR1(form = ~ Year), knots = knots)
}

## generate new trends using draws from age-model posterior
simTrendMods <- lapply(ageSims %>% split(ageSims$draw), fitSWModels,
    y = small$d15N, knots = knots)

@@

```{r, dependson = -1}
## wrapper to simulate from a fitted GAM with the required arguments
simulateSWModels <- function(mod, newdata, nsim, seed = 42) {
    sims <- fitted_samples(mod, n = nsim, data = newdata, seed = seed)
    pull(sims, .fitted)
}

@@

```{r small-scam-fit-plots, dependson = -1}
## plot the estimated age model ad posterior simulations from it
plt1 <- ggplot(swAge, aes(y = Date, x = Depth)) +
    geom_line(data = newSims,
              mapping = aes(y = value, x = Depth, group = draw),
              alpha = 1, colour = "grey80") +
    geom_line(data = newAgeFit, mapping = aes(y = .fitted, x = Depth)) +
    geom_point(size = 1.5, colour = "red") +
    geom_errorbar(aes(ymin = Date - Error, ymax = Date + Error, width = 0),
                  colour = "red") +
    labs(y = "Year CE", x = "Depth (cm)")
